Defining properties regulating pioneer-factor binding and function

### 2023-02-03 manuscript/manuscript_sections/_04_Discussion.qmd
update MS text

Major conclusions to emphasize:

-   Twist is a pioneer factor

-   PF activity occurs on a spectrum and is context-dependent.

-   Context-dependent PF activity can be regulated by both protein-intrinsic and extrinsic features.

-   Preexisting active chromatin attracts TF binding, even with weaker motifs.

-   Most functional binding occurs at canonical motifs in naïve chromatin

-   Sequence composition is a main determinant of PF binding and activity

-   Regions outside the DBD are essential for pioneering

>>>>

## outline:

### Summary of study

### Pioneer-factor activity occurs on a spectrum and is context dependent

-   Twi is a PF

-   In our study, we see a range of pioneering activities: Twi\<Zld\<Grh

### Pioneer factors are regulated by both protein-intrinsic and protein-extrinsic features Local chromatin.

-   **Active chromatin:**

    -   promotes PF binding in the absence fo strong motifs.

    -   H3K27me3, despite correlation, does not affect binding and pioneering

-   **Protein concentration**:

    -   Strong determinant of binding and chromatin opening in our study.

    -   Different binding sites have different thresholds for binding and opening.
        Some remain unbound even at high levels.

-   **Cofactors**?

-   Not explored extensively in our study.
    How to discuss without inviting more experiments from reviewers?

-   **Motif content**:

    -   Strong predictor of binding and opening.

-   **Protein domains**:

    -   Required for stable binding and opening.

    -   Allows uncoupling of binding from opening.

### Model:

-   PF activity requires high chromatin occupancy which can be achieved in a variety of ways.

### Regions outside the DBD are essential for pioneering

-   This and other recent evidence complicates the simple modular model of TF function

-   Multiple possibilities for how IDRs or activation domains could impact pioneering:

    -   multivalent interactions with nucleosomes stabilize binding (supported by in vitro evidence).

    -   Recruitment/interaction with cofactors that stabilizes binding (chromatin opening via cofactors could provide positive feedback that stabilizes binding)

    -   Allowing broader exploration of nuclear space, including repressed domains (supported by imaging data)

    -   Clustering or phase separation that increases local concentration

### Future directions we could mention:

-   How do Zld, Grh, Twi and other PFs interact with nucleosomal targets?
    TFs likely have diverse mechanisms of interaction, making it important to characterize for many factors.

-   What are the mechanisms of chromatin opening for these factors?
    Do they recruit remodelers?
    Does binding alone destabilize nucleosomes?

-   How do regions outside the DBD contribute to binding and opening?

### 2023-02-03 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update MS text
update outline of supplemental figures

**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations.**

-   ~~Schematic of cell line generation.~~

-   ~~Western blots to determine physiological expression and lack of protein expression in WT cells.~~

-   ~~RNA-seq data to show leaky expression and strong induction of RNA levels.~~

@@

@@

**Additional information to include in supplemental figures:**

-   De novo motif results

-   Analysis of promoter motifs

-   Analysis of deep learning models

>>>>

**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations. a,** Schematic of generation of stable cell lines and induction of protein expression.
**b,** mRNA levels of Zld and Grh in S2 cells.
Top histogram shows the distribution of mRNA levels for all Drosophila genes.
Vertical dashed line indicates a log~2~ RPKM value of 0 as a threshold for considering a gene to be expressed.
**c-d,** Immunoblots showing titration of Zld **(c)** or Grh **(d)** protein levels in stable cell lines.
Two independently generated cell lines are shown and compared to 2-3H embryos.

@@

-   Western blots for determining DBD expression levels

-   IF to confirm nuclear localization of DBDs.

@@
